Automatic Identification and Annotation of Tasks in Structured Programs

### 2018-01-28 docs/ACM/paper.tex
Adding discussion about windmills.

% Windmills and vanes
A {\em task candidate} is a program region that can run in parallel with the
rest of the program.
To identify task candidates, we rely on the structure of the program's
control flow graph.
To this end, we identify, subgraphs in the CFG called {\em windmills}, a term
coined by Rideau {\em et al.}~\cite{Rideau08} to describe structural relations
between register copies.
We adopt a slightly more general definition than Rideaus'; however, the
graphical shape of our graphs remain similar:

\begin{definition}[Windmill]
\label{def:windmill}
A windmill is a graph $G_w = G_c \cup G_{v1} \cup \ldots \cup G_{vn}$ formed by
a strongly connected component $G_c$, its center, plus $n$ components (not
necessarily strong) $G_{vi}$, the vanes, such that:
\begin{compactenum}
\item for any topological ordering of $G_w$, and nodes $n_c \in G_c$,
$n_v \in G_{vi}$, we have $n_c$ ahead of $n_v$;
\item for any $i$ and $j$, $1 \leq i < j \leq n$, $G_{vi}$ and $G_{vj}$ do not
share vertices.
\end{compactenum}
\end{definition}

Statically, a strongly connected set $S$ in the  \emph{Dependence Graph} $G$ represents cyclic dependence between instructions in any iterative construct.
It arises from a loop in the program or, if $G$ is interprocedural,
from a recursive pattern.

Let $L = \{S_{1}, S_{2}, ..., S_{n}\} \subset W$ be the list of $n$ strongly connected sets contained
in a windmill $W$. We call $S_{b}$ to be a \emph{windmill base} if $\forall S_{i} \subset L$, $S_{b} \geq S_{i}$ in any topological sorting of $G$.

\begin{definition}[Helix]
\label{def:helix}
Given a windmill $W$  and a windmill base $S \subset W$ in the \emph{Dependence Graph} $G$, a set of nodes $H$ is a Helix if, and only if:
\begin{itemize}
\item $H \subset W$;
\item $H  \cap S_{b} = \emptyset$;
\item $H$ is connected and maximal.
\end{itemize}
\end{definition}

We can view a \emph{Helix} as a set of nodes in the \emph{Dependence Graph}
that depends on a strongly connected set which is a \emph{windmill base}.
That means helices are all those subgraphs that depend on
the \emph{windmill base} but do not form a cyclic dependence with it.
Figure \ref{fig:helix} illustrates both concepts.

The insight behind  \emph{Helices} is that they represent instructions that 
are deemed to be executed more than once, for they depend on a cyclic
dependence (\emph{windmill base}), and, since they do not form a strongly
connected set itself, they can be executed in parallel.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/helix}
\caption{\emph{Windmills} and \emph{Helices} corresponding to the sample code above.
We see the \emph{windmill} 
$W = \{A, B, C, D, E, F, G, H\}$, the windmill base $S_{b} = \{A, B, C\}$ 
and two \emph{helices} $H_{1} = \{H\}$ and $H_{2} = \{D, E, F, G\}$.  Note
that there are more than one strongly connected, but only of them is a
base, which is the one at the top topologically.}
\label{fig:helix}
\end{center}
\end{figure}

Thus, the first step in the algorithm to find tasks is to map these \emph{Helices}
into program regions. The algorithm scans through the \emph{Dependence Graph}
looking for \emph{Helices} and then finds the minimal covering region for each \emph{Helix}.
These minimal regions are later going to be annotated as tasks in the original
program, should they fit a cost model.

After finding potential task regions, the next critical step is the \emph{expansion}.

@@

\pedro{I asked gleison and he'll sit with me to explain everything ,
how it's done, and then I'll write it here.}

>>>>

% Windmills and vanes
A {\em task candidate} is a set of program statements that can run in parallel
with the rest of the program.
To identify task candidates, we rely on the program's
{\em Dependence Graph}.
For completeness, Definition~\ref{def:pdg} revisits this notion.

\begin{definition}[Program Dependence Graph -- \cite{Ferrante87}]
\label{def:pdg}
Given a program $P$, its PDG contains one vertex for each statement $s \in P$.
There exists an edge from $s_1$ to $s_2$ if the latter depends on the former.
Statement $s_2$ is {\em data-dependent} on $s_1$ if it reads data that $s_1$
writes.
It is {\em control dependent} on $s_1$ if $s_1$ determines the outcome of a
branch, and depending on this outcome, we might execute $s_2$ or not.
\end{definition}

Task candidates are vanes in windmills.
{\em Windmills} are a family of graphs.
To the best of our knowledge, the term was coined by Rideau
{\em et al.}~\cite{Rideau08} to describe structural relations between register
copies.
Our windmills exist as subgraphs of the program's dependence graph.
We adopt a slightly more general definition than Rideaus'; however, the
metaphor that gave origin to the name: the shape of the graphs, is still
unmistakable:

\begin{definition}[Windmill -- \cite{Rideau08}]
\label{def:windmill}
A windmill is a graph $G_w = G_c \cup G_{v1} \cup \ldots \cup G_{vn}$ formed by
a strongly connected component $G_c$, its center, plus $n$ components (not
necessarily strong) $G_{vi}$, the vanes, such that:
\begin{compactenum}
\item for any topological ordering of $G_w$, and nodes $n_c \in G_c$,
$n_v \in G_{vi}$, we have $n_c$ ahead of $n_v$;
\item for any $i$ and $j$, $1 \leq i < j \leq n$, $G_{vi}$ and $G_{vj}$ do not
share vertices.
\end{compactenum}
\end{definition}

\begin{figure}[h]
\begin{center}
\includegraphics[width=1\columnwidth]{images/ex_windmill}
\caption{Examples of program dependence graphs.
Nodes that form centers of windmills are coloured in gray.
Data-dependence edges are solid; control-dependence edges are dashed.
Vanes of windmills appear within dotted boxes.
(a) PDG for Figure~\ref{fig:ex_book_filter};
(b) PDG for Figure~\ref{fig:ex_Regions}}
\label{fig:ex_windmill}
\end{center}
\end{figure}

Figure~\ref{fig:ex_windmill} shows two program dependence graphs, highlighting
windmills and their vanes.
The structure of windmills naturally leads to task candidates.
Vanes correspond to program parts that are likely to be executed
several times, because they sprout from a loop -- the center of the windmill.
Two different executions of the same vane can run in parallel.
This is a property of the PDG, and one of the original motivations behind its
design.
These two different executions can be seen as two instances of the same
structure (the vane), coming out of the same windmill center.
A topological ordering of the dependence graph does not impose any order among
these two hypothetical replicas of the same structure.

Windmills gives us structural properties to identify tasks.
We can easily find them via a depth-search traversal of the program dependence
graph, for instance.
However, not every windmill lead us to profitable tasks.
Furthermore, some windmills simply cannot be annotated, due to structural 
incompatibilities between the program dependence graph and its source code.
We deal with the first shortcoming in Section~\ref{sub:profit}; the latter is
the subject of Section~\ref{sub:ir}.

@@
